Are Transformed Low Discrepancy Points Also Low Discrepancy?

### 2019-05-19 InvPts_verB.tex
latest updates

To illustrate this point, consider a cousin of the kernel in \eqref{eq:OrigKernel}, which places the reference point at $\boldsymbol{0.5} = (0.5, \ldots, 0.5)$, the center of the unit cube $(0,1)^d$:
\begin{align}
\label{eq:L2kernel}
    K_{\unif}(\vu,\vv) &= \prod_{j=1}^d\left[1+\frac{1}{2}|u_j-0.5|+\frac{1}{2}|v_j-0.5|-\frac{1}{2}|u_j-v_j|\right] \\
    \nonumber
    & = K(\vu - \boldsymbol{0.5}, \vv - \boldsymbol{0.5}) \qquad \text{for $K$ defined in \eqref{eq:OrigKernel}}. 
\end{align}
This kernel defines the centered $L^2$-discrepancy \cite{Hic97a}.
Consider the standard multivariate normal distribution, $F_\normal$, and choose the inverse normal distribution,
\begin{equation*}
    \vPsi(\vu) = (\Phi^{-1}(u_1), \ldots, \Phi^{-1}(u_d)),
\end{equation*}
where $\Phi$ denotes the standard normal distribution function.  Then condition \eqref{eq:BigQYesb} is automatically satisfied, and condition \eqref{eq:BigQYesa} is  satisfied by defining
\begin{align}
\nonumber
     K(\vt,\vx) &= K_{\unif}(\vPsi^{-1}(\vt),\vPsi^{-1}(\vx)) \\
     \nonumber
      &= \prod_{j=1}^d\left[1+\frac{1}{2}|\Phi^{-1}(t_j)-0.5|+\frac{1}{2}|\Phi^{-1}(x_j)-0.5| \right . \\
      & \qquad \qquad \left . -\frac{1}{2}|\Phi^{-1}(t_j)-\Phi^{-1}(x_j)|\right].
\end{align}
In one dimension, the distance between two Dirac measures defined using the kernel $K_\unif$ above is $\norm{\delta_{x} - \delta_{t}} = \sqrt{x-t}$, whereas the distance defined using the kernel $K$ above is $\norm{\delta_{x} - \delta_{t}} = \sqrt{\Phi^{-1}(x)-\Phi^{-1}(t)}$.  Under kernel $K$, the distance between two Dirac measures can never be greater than one.  Such an assumption may be unpalatable.

@@

From the discussion above, it appears that condition \eqref{eq:BigQYesa} can be too restrictive.  We would like to compare the discrepancies of designs under kernels that do not satisfy that restriction.  In particular, we consider the centered $L^2$-discrepancy for designs on $(0,1)^d$ defined by the kernel in \eqref{eq:L2kernel}:
\begin{align*}
\MoveEqLeft{D^2(\Udes, F_\unif, K_\unif)} \\
& = \left(\frac{13}{12}\right)^d - \frac{2}{N}\sum_{i=1}^N \prod_{j=1}^d \left(1+\frac{1}{2}|u_{ij}-1/2|-\frac12|u_{ij}-1/2|^2\right)\\
& \qquad + \frac{1}{N^2}\sum_{i,k=1}^N\prod_{j=1}^d\left[1+\frac{1}{2}|u_{ij}-1/2|+\frac{1}{2}|u_{kj}-1/2|-\frac{1}{2}|u_{ij}-u_{kj}|\right],    
\end{align*}
We also consider the discrepancy for designs defined with respect to the standard normal distribution defined by the kernel \eqref{eq:OrigKernel}.  We choose this kernel because it has a more realistic interpretation in terms of the distance between two Dirac measures.  This discrepancy is 
\begin{align}\label{eq:DisNormald>1}
\nonumber
\MoveEqLeft{D^2(\Xdes, F_\normal, K) = \left(1+\sqrt{\frac{2}{\pi}}\right)^d}\\
\nonumber
  & - \frac{2}{N}\sum\limits_{\vx\in \Xdes} \prod\limits_{j=1}^d\left[ 1+\frac{1}{\sqrt{2\pi}}+\frac{1}{2}|x_j|-x_j[\Phi(x_j)-1/2]-\phi(x_j)\right]\\
  &+\frac{1}{N^2}\sum_{\vx,\vt\in \Xdes}\prod_{j=1}^d \left[1+\frac{1}{2}|x_j|+\frac{1}{2}|t_j|-\frac{1}{2}|x_j-t_j|\right]. 
\end{align}
When $d=1$, the discrepancy \eqref{eq:DisNormald>1} can be further simplified to
\begin{align}\label{eq:DisNormald=1}\nonumber
D^2(\Xdes, \Phi_1, K)=& -\frac{2}{N}\sum_{i=1}^N \left[\max(x_i,0)-x_i\Phi_1(x_i)-\phi(x_i)\right]\\
+&\frac{1}{N}\sum_{i=1}^N|x_i|-\frac{1}{2N^2}\sum_{i,j=1}^N|x_i-x_j|. 
\end{align}
In both \eqref{eq:DisNormald=1} and \eqref{eq:DisNormald>1}, $\phi(\cdot)$ is the pdf of $N(0,1)$. 
The derivations are in the Appendix.

The discrepancy for the uniform distribution on unit cube $(0,1)^d$ we adapt in this work is the centered $L_2$-discrepancy \cite{Hic97a} defined as:
\begin{align*}
[D_2(\Udes, F_\unif, K_2)]^2 =& \left(\frac{13}{12}\right)^d - \frac{2}{N}\sum\limits_{\vu\in \Udes}\prod_{j=1}^d \left(1+\frac{1}{2}|u_j-1/2|-1/2|u_j-1/2|^2\right)\\
\hspace{5ex}&+ \frac{1}{N^2}\sum\limits_{\vu,\vu'\in \Udes}\prod_{j=1}^d\left[1+\frac{1}{2}|u_j-1/2|+\frac{1}{2}|u_j'-1/2|-\frac{1}{2}|u_j-u_j'|\right],    
\end{align*}
with the reproducing kernel:
\begin{equation}\label{eqn:oril2kernel}
K_2(\vx,\vt) = \prod\limits_{j=1}^d\left[1+\frac{1}{2}|x_j-1/2|+\frac{1}{2}|t_j-1/2|-\frac{1}{2}|x_j-t_j|\right],
\end{equation}
which is the same as \eqref{eq:OrigKernel}, except it is anchored to the center $(1/2,\cdots,1/2)$ of the cube but not the origin. The centered $L_2$-discrepancy is invariant under reflections of $\Udes$ about any plane $x_j = 1/2$.

With the kernel in \eqref{eq:OrigKernel}, now we define a discrepancy for multivariate standard normal distribution on $\mathbb{R}^d$ that also refers to the center of the experimental region, $(0,\cdots,0)$. Firstly, we show that this kernel function is a reproducing kernel with a properly defined inner product as follows. 
\begin{equation}\label{eq:inner}
\langle f,g \rangle = \sum_{u\subseteq \{1,...,d\}}\int_{\mathbb{R}^{|u|}}\frac{\partial^{|u|}}{\partial \vx_u}f(\vx_u,{\bf 0})\frac{\partial^{|u|}}{\partial \vx_u}g(\vx_u,{\bf 0})\dif \vx_u,
\end{equation}
where $u$ is an index set and $u\subseteq \{1, \ldots, d\}$ and $|u|$ is the cardinality of $u$. 
Using the following operators, 
$$\partial \vx_u = \prod\limits_{k\in u}\partial x_k \text{ and } \dif \vx_u = \prod\limits_{k\in u}\dif x_k , $$
the $|u|$-th order partial derivative in \eqref{eq:inner} is with respective to $\vx_{u}$, the subset of variables indexed by $u$, and the integration is also with respect to the same subset of variables. 
The vector $(\vx_u,{\bf 0})$ is a $d$-dimensional vector with the dimension indexed by $u$ denoted as $\vx_u$ and the rest are set to be ${\bf 0}$. 
In particular, for $u = \{1,...,d\}$, we have $(\vx_u,{\bf 0}) = \vx$; for $u=\emptyset$, we have $(\vx_u,{\bf 0}) = {\bf 0}$. In particular, when $d=1$,
\begin{equation*}
\langle f,g \rangle = f(0)g(0)+\int_{\mathbb{R}}f'(x)g'(x)\dif x.
\end{equation*}

With the inner product \eqref{eq:inner}, it is easy to see that above kernel function in \eqref{eq:OrigKernel} is symmetric and positive semi-definite on $\Omega = \mathbb{R}^d$, i.e., the domain of the $d$-dim multivariate normal distribution. 
Note that $K({\bf 0},\vt) = 1$ and
\begin{align*}
\frac{\partial^{|u|} K((\vx_u,{\bf 0}),\vt)}{\partial \vx_u}& =\prod_{i\in u}\frac 12 \left[\sign(x_i) - \sign(x_i - t_i) \right] \\
= &\begin{cases}
\prod\limits_{i\in u}\sign(t_i), & \min(0,t_i) < x_i \leq \max(0,t_i), i\in u\\
0, & \textrm{otherwise}.
\end{cases}.
\end{align*}
Thus, $\langle K(\cdot,\vt), K(\cdot,\vt) \rangle < \infty $, and
\begin{align*}
\langle f, K(\cdot,\vt) \rangle &= \sum_{u\subseteq \{1,...,d\}}\int_{\mathbb{R}^{|u|}}\frac{\partial^{|u|}}{\partial \vx_u}f(\vx_u,{\bf 0})\frac{\partial^{|u|} K((\vx_u,{\bf 0}),\vt)}{\partial \vx_u}\dif \vx_u \\
& =     \sum_{u\subseteq \{1,...,d\}}\sum_{v\subseteq u}(-1)^{|u|-|v|}f(\vt_v,{\bf 0})\\
& = f(\vt)
\end{align*}
With the reproducing kernel \eqref{eq:OrigKernel}, we can analytically derive the discrepancy for the standard normal distribution.

Let $\Xdes = \{\vx_i\}_{i=1}^N$ be a design on $\Omega = \mathbb{R}^d$. 
The squared discrepancy of the design $\Xdes$ with respect to the $d$-dimensional standard multivariate normal probability distribution $N({\bf 0}, {\bf I}_d)$ with cdf $\Phi$ is
\begin{eqnarray}\label{eqn:discrepancy}\nonumber
D^2(\Xdes, \Phi, K) &=& \int_{\Omega\times\Omega} K(\vx,\vt)\dif \Phi(\vx)\dif \Phi(\vt)\\
&-&\frac{2}{N}\sum_{i=1}^N \int_{\Omega}K(\vx,\vx_i)\dif \Phi(\vx)+\frac{1}{N^2}\sum_{i,k=1}^N K(\vx_i,\vx_k).
\end{eqnarray}
The cdf $\Phi(\vx)=\prod_{j=1}^d \Phi_1(x_j)$, where $\Phi_1(\cdot)$ is the cdf of $N(0,1)$. 
With simple calculation, the squared discrepancy \eqref{eqn:discrepancy} can be  computed explicitly as
\begin{align}\label{eq:DisNormald>1}\nonumber
&D^2(\Xdes, \Phi, K)\\\nonumber
  &= \left(1+\sqrt{\frac{2}{\pi}}\right)^d - \frac{2}{N}\sum\limits_{\vx\in \Xdes} \prod\limits_{j=1}^d\left[ 1+\frac{1}{\sqrt{2\pi}}+\frac{1}{2}|x_j|-x_j[\Phi_1(x_j)-1/2]-\phi(x_j)\right]\\
  &+\frac{1}{N^2}\sum_{\vx,\vt\in \Xdes}\prod_{j=1}^d \left[1+\frac{1}{2}|x_j|+\frac{1}{2}|t_j|-\frac{1}{2}|x_j-t_j|\right]. 
\end{align}
When $d=1$, the discrepancy \eqref{eq:DisNormald>1} can be further simplified to
\begin{align}\label{eq:DisNormald=1}\nonumber
D^2(\Xdes, \Phi_1, K)=& -\frac{2}{N}\sum_{i=1}^N \left[\max(x_i,0)-x_i\Phi_1(x_i)-\phi(x_i)\right]\\
+&\frac{1}{N}\sum_{i=1}^N|x_i|-\frac{1}{2N^2}\sum_{i,j=1}^N|x_i-x_j|. 
\end{align}
In both \eqref{eq:DisNormald=1} and \eqref{eq:DisNormald>1}, $\phi(\cdot)$ is the pdf of $N(0,1)$. 
The derivations are in the Appendix.

\section{Inconsistency in Discrepancy}\label{sec:inconsistency}
Using the inverse transformation method, we can generate the standard multivariate normal distribution $N({\bf 0},{\bf I}_d)$ by applying $x_{i,j}=\Phi^{-1}_j(u_{i,j})$ for $j=1,\ldots,d$ and $i=1,\ldots, N$. 
However, the inverse transformation would lead to inconsistency between the discrepancy before and after the transformation. 
Although $\Udes = \{\vu_i\}_{i=1}^N$ is the low-discrepancy points for the uniform distribution in $[0,1]^d$, the following examples shows that it is not necessarily so that $\Xdes=\{\vx_i\}_{i=1}^N$ is also low-discrepancy. 
In other words, for two uniform designs, $\Udes_1$ and $\Udes_2$, the fact that $D(\Udes_1,F_\unif, K)\leq D(\Udes_2,F_\unif, K)$ does not imply that the inverse transformed normal design $\Xdes_1$ and $\Xdes_2$ also have $D(\Xdes_1,\Phi, K)\leq D(\Xdes_2,\Phi, K)$. 
Note that the two discrepancies $D(\Udes, F_\unif, K)$ and $D(\Xdes, \Phi, K)$ are based on the same kernel, hence the inconsistency is only due to the inverse transformation.

Let $d=5$ and $n=50$. 
We generate $B=20$ samples from uniform distribution in $[0,1]^5$ and then use the inverse distribution transformation to obtain samples following $N({\bf 0}, {\bf I}_5)$. 
The discrepancy for uniform distribution is the $L_2$-discrepancy $D_2(\Udes,F_\unif,K_2)$ and for the standard normal distribution is \eqref{eq:DisNormald>1}. 
Figure \ref{fig:UniVsNormDisc} plots the discrepancy for normal distribution against the $L_2$-discrepancy for each of the $B=20$ samples. 
If the inverse transformation makes the design consistent in terms of the discrepancy, the points in \ref{fig:UniVsNormDisc} should be approximately aligned on a straight line. 
Clearly, some samples with low uniform discrepancy turn out to have large normal discrepancy after inverse transformation.
Thus, inverse transformation cannot ensure the consistency of the design in terms of the discrepancy.

>>>>

To illustrate this point, consider a cousin of the kernel in \eqref{eq:OrigKernel}, which places the reference point at $\boldsymbol{0.5} = (0.5, \ldots, 0.5)$, the center of the unit cube $(0,1)^d$:
\begin{align}
\label{eq:L2kernel}
    K_{\unif}(\vu,\vv) &= \prod_{j=1}^d\left[1+\frac{1}{2}\left(\left|u_j-1/2\right|+ \left|v_j- 1/2 \right|-\left |u_j-v_j \right| \right) \right] \\
    \nonumber
    & = K(\vu - \boldsymbol{0.5}, \vv - \boldsymbol{0.5}) \qquad \text{for $K$ defined in \eqref{eq:OrigKernel}}. 
\end{align}
This kernel defines the centered $L^2$-discrepancy \cite{Hic97a}.
Consider the standard multivariate normal distribution, $F_\normal$, and choose the inverse normal distribution,
\begin{equation} \label{eq:invNormTrans}
    \vPsi(\vu) = (\Phi^{-1}(u_1), \ldots, \Phi^{-1}(u_d)),
\end{equation}
where $\Phi$ denotes the standard normal distribution function.  Then condition \eqref{eq:BigQYesb} is automatically satisfied, and condition \eqref{eq:BigQYesa} is  satisfied by defining
\begin{align}
\nonumber
     K(\vt,\vx) &= K_{\unif}(\vPsi^{-1}(\vt),\vPsi^{-1}(\vx)) \\
     \nonumber
      &= \prod_{j=1}^d\left[1+\frac{1}{2}\left(\left|\Phi^{-1}(t_j)-1/2\right|+ \left|\Phi^{-1}(x_j)- 1/2 \right| \right . \right . \\
      & \qquad \qquad \left.  \left . -\left|\Phi^{-1}(t_j)-\Phi^{-1}(x_j)\right|\right)\right].
\end{align}
In one dimension, the distance between two Dirac measures defined using the kernel $K_\unif$ above is $\norm{\delta_{x} - \delta_{t}} = \sqrt{x-t}$, whereas the distance defined using the kernel $K$ above is $\norm{\delta_{x} - \delta_{t}} = \sqrt{\Phi^{-1}(x)-\Phi^{-1}(t)}$.  Under kernel $K$, the distance between two Dirac measures can never be greater than one.  Such an assumption may be unpalatable.

@@

From the discussion above indicates that condition \eqref{eq:BigQYesa} can be too restrictive.  We would like to compare the discrepancies of designs under kernels that do not satisfy that restriction.  In particular, we consider the centered $L^2$-discrepancy for uniform designs on $(0,1)^d$ defined by the kernel in \eqref{eq:L2kernel}:
\begin{align*}
\MoveEqLeft{D^2(\Udes, F_\unif, K_\unif)} \\
& = \left(\frac{13}{12}\right)^d - \frac{2}{N}\sum_{i=1}^N \prod_{j=1}^d \left[1+\frac{1}{2}\left(|u_{ij}-1/2|-|u_{ij}-1/2|^2\right) \right]\\
& \qquad + \frac{1}{N^2}\sum_{i,k=1}^N\prod_{j=1}^d\left[1+\frac{1}{2}\left(|u_{ij}-1/2|+|u_{kj}-1/2|-|u_{ij}-u_{kj}| \right) \right],  
\end{align*}
where again, $F_\unif$ denotes the uniform distribution on $(0,1)^d$, and $\Udes$ denotes a design on $(0,1)^d$

Changing perspectives slightly, if $F_\unif'$ denotes the uniform distribution on the cube of volume one centered at the origin, $(-\boldsymbol{0.5},\boldsymbol{0.5})^d$, and the design $\Udes'$ is constructed by  subtracting $\boldsymbol{0.5}$ from each point in the design $\Udes$:
\begin{equation} \label{eq:Updef}
    \Udes' = \{\vu -\boldsymbol{0.5} : \vu \in \Udes \},
\end{equation}
then 
\begin{equation} \label{eq:desEquiv}
     D(\Udes',F'_\unif,K) = D(\Udes, F_\unif, K_\unif),
\end{equation}
where $K$ is the kernel defined above in \eqref{eq:OrigKernel}.  Recall that the origin is a special point in the definition of the inner product corresponding to this $K$ as a reproducing kernel in \eqref{eq:inner}.

Therefore, this $K$ from \eqref{eq:OrigKernel} is an appropriate kernel to use to define the the discrepancy for designs defined with respect to the standard normal distribution, which is also centered at the origin.  Such a discrepancy is 
\begin{align}\label{eq:DisNormald>1}
\nonumber
\MoveEqLeft{D^2(\Xdes, F_\normal, K) = \left(1+\sqrt{\frac{2}{\pi}}\right)^d}\\
\nonumber
  & - \frac{2}{N}\sum_{i=1}^N \prod\limits_{j=1}^d\left[ 1+\frac{1}{\sqrt{2\pi}}+\frac{1}{2}|x_{ij}|-x_{ij}\left(\Phi(x_{ij})-\frac 12 \right)-\phi(x_{ij})\right]\\
  &+\frac{1}{N^2}\sum_{i,k=1}^N \prod_{j=1}^d \left[1+\frac{1}{2}\left(|x_{ij}|+|x_{kj}|-|x_{ij}-x_{kj}|\right)\right]. 
\end{align}
Here, $\phi$ is the standard normal probability density function. 
The derivation of \eqref{eq:DisNormald>1} is given in the Appendix.

%\section{Inconsistency in Discrepancy}\label{sec:inconsistency}
Now we numerically compare the discrepancy of a uniform design, $\Udes'$ given by \eqref{eq:Updef} and the discrepancy of a design constructed by the inverse normal transformation, i.e.,  $\Xdes = \vPsi(\Udes)$ for $\vPsi$ in \eqref{eq:invNormTrans}, where the $\Udes$ leading to both is identical.  We do not expect the magnitudes of the discrepancies to be the same, but we ask
\begin{multline} \tag{Q'}
    \text{does } D(\Udes'_1,F'_\unif, K)\leq D(\Udes_2',F_\unif', K) \\
    \text{imply }
     D(\vPsi(\Udes_1),F_\normal, K)\leq D(\vPsi(\Udes_2),F_\normal, K)?
\end{multline}
Again, $K$ is given by \eqref{eq:OrigKernel}.  So we are actually comparing discrepancies defined by the same kernels, but \emph{not kernels that satisfy \eqref{eq:BigQYesa}}.

\FJH{Fred got to here}

Let $d=5$ and $N=50$. 
We generate $B=20$ independent and identically distributed (IID) uniform designs with $50$ points on $(0,1)^5$ and then use the inverse distribution transformation to obtain samples following $N({\bf 0}, {\bf I}_5)$. 
The discrepancy for uniform distribution is the $L_2$-discrepancy $D_2(\Udes,F_\unif,K_2)$ and for the standard normal distribution is \eqref{eq:DisNormald>1}. 
Figure \ref{fig:UniVsNormDisc} plots the discrepancy for normal distribution against the $L_2$-discrepancy for each of the $B=20$ samples. 
If the inverse transformation makes the design consistent in terms of the discrepancy, the points in \ref{fig:UniVsNormDisc} should be approximately aligned on a straight line. 
Clearly, some samples with low uniform discrepancy turn out to have large normal discrepancy after inverse transformation.
Thus, inverse transformation cannot ensure the consistency of the design in terms of the discrepancy.
